---
title: "Salivary Proteomics in Sheep"
author: "Jared Raynes"
date: "27 April 2018"
output: 
  html_document:
    keep_md: true
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(lmerTest)
library(emmeans)
library(factoextra)
```

## Background

R scripts and markdown documents can be found at <https://github.com/jaredraynes/Sheep_Saliva>

Background and data provided by Dr Michelle Colgrave

Dark cutting is a notoriously difficult trait to assess. In this trial we proposed to assess dark cutting by a combination of muscle glycogen content and pH 24 hours post-slaughter. We also performed quantitative proteomics analysis on the saliva taken from sheep under different conditions of nutrition.

### Experimental design

Sixty wethers (castrated males) of around 1 month of age were brought to the Animal House, and allocated into 4 groups (n=15 per group) balanced by weight. Groups 1 and 2 were allocated to High Nutrition treatment, while groups 3 and 4 were allocated to Low Nutrition treatment. The trial had four treatment groups (2 x 2 factorial design):   
1)	ad lib feed; no stress  
2)	ad lib feed: stress*  
3)	restricted feed (sloping 70 – 60 % of maintenance); no stress  
4)	restricted feed; stress* 

Note: the stress event occurred after the saliva was taken and as such should not impact these results.

The sheep were assessed for glycogen content and the pH of the meat was measured 24h after slaughter.

The pH was assessed after 24 h at 4 degress (final pH). A final pH > 5.7 is considered a cause for downgrading in Meat Standards Australia, and seen as indicative of ‘dark cutting’. Note that pH in the leg muscle and in the back muscle are quite different. The leg muscle (M. semitendinosus) is particularly sensitive to glycogen depletion.

### Data provided

The data was provided in the form of two Microsoft Excel documents.  
1. "Copy of 20180413_DataSchool_Glycogen.xlsx"  
2. "Copy of Data Matrix_Peptide Peak Area.xlsx"

#### Initial investigation of Excel 1

```{r sheep}
excel_sheets("Data/Copy of 20180413_DataSchool_Glycogen.xlsx")
```

####Contains 5 sheets within the Excel document

## pH Data

####Inital look at data
```{r pH}
pH <- read_xlsx("Data/Copy of 20180413_DataSchool_Glycogen.xlsx", sheet = "pH")
pH
```
####Tidying the data

```{r pH_tidying}
pH$Brand[13] = 64 #replacing data from sheep 15 with 64
pH <- pH[1:4]
names(pH) <- c("Treatment.Group", "ID", "M.semitendinosus", "M.longissimus.dorsi")
pH <- mutate(pH, Feed = ifelse(Treatment.Group <= 2, TRUE, FALSE)) %>%  #added feed column
  mutate(Stress = ifelse(Treatment.Group == 2|Treatment.Group == 4, TRUE, FALSE))
pH_long <- pH %>% 
  gather(key = "Muscle.type", value = "pH", "M.semitendinosus", "M.longissimus.dorsi")
pH_long[1:5] <- lapply(pH_long[1:5], as.factor)
pH_long
```

#### Initial visulisation of data

```{r pH_plot1, echo=FALSE}
ggplot(pH_long, aes(x = Treatment.Group, y = pH, colour = Muscle.type))+
  geom_boxplot()+
  labs(x = "Treatment Group", y = "pH") +
  theme(axis.title.y = element_text(size = rel(1.6), angle = 90)) +
  theme(axis.title.x = element_text(size = rel(1.6), angle = 00))

ggplot(pH_long, aes(x = Stress, y = pH, colour = Feed))+
  geom_jitter() +
  facet_wrap(~Muscle.type) +
  labs(x = "Stress", y = "pH") +
  theme(axis.title.y = element_text(size = rel(1.6), angle = 90)) +
  theme(axis.title.x = element_text(size = rel(1.6), angle = 00))
```

#### Modelling of data

```{r pH_modelling}
lm.pH1 <- lm(pH ~ Stress*Feed + Muscle.type, data = pH_long)
summary(lm.pH1)
emmeans(lm.pH1, pairwise~Stress+Feed)
plot(lm.pH1, which =1)
```

#### Modelling Results
1. Feeding regime only had a significant effect (P = 0.005) on the pH of the M.semitendinosus muscle and this only occurred when the animals were stressed.  
2. Stress had a significant (P < 2e-16) impact on muscle pH.  
3. The M.semitendinosus muscle had a significantly (P = 1.91e-12) higher pH compared to the M.longissimus.dorsi muscle. 
4. None of the treatments resulted in 'dark cutting' of meat.

## Glycogen Data (Taken from M.semitendinosus Muscle)

####Inital look at data
```{r Glycogen}
Glycogen <- read_xlsx("Data/Copy of 20180413_DataSchool_Glycogen.xlsx", sheet = "Glycogen")
Glycogen
```
####Tidying the data

```{r Glycogen_tidying}
Glycogen <- Glycogen[-c(31),] ## remove error message
G1 <- Glycogen[1:5] #the treatment groups are in different columns this time
G2 <- Glycogen[6:10] #need to have identical names to use row_bind
names(G2) <- c("Treatment group", "Brand", "Glycogen mg/g tissue", "Lactate mg/g", "Corrected glycogen")
G2$`Treatment group` <- as.character(G2$`Treatment group`) 
Glycogen_comb <- bind_rows(G1, G2) #combining the columns together
Glycogen_comb <- mutate(Glycogen_comb, Feed = ifelse(`Treatment group` <= 2, TRUE, FALSE)) %>%  #added feed column
  mutate(Stress = ifelse(`Treatment group` == 2|`Treatment group` == 4, TRUE, FALSE)) #added Stress column
Glycogen_long <- Glycogen_comb %>% 
  gather(key = "Measure", value = "mg/g", "Glycogen mg/g tissue", "Lactate mg/g", "Corrected glycogen")
names(Glycogen_long) <- c("Treatment.Group", "ID", "Feed", "Stress", "Measure", "mg/g")
Glycogen_long[1:5] <- lapply(Glycogen_long[1:5], as.factor)
Glycogen_sel <- filter(Glycogen_long, Measure == "Corrected glycogen")
Glycogen_long
```

#### Initial visulisation of data

```{r Glygogen_plot1, echo=FALSE}
ggplot(Glycogen_long, aes(x = Treatment.Group, y = `mg/g`, colour = Measure))+
  geom_boxplot()+
  labs(x = "Treatment Group", y = "Glycogen mg/g") +
  theme(axis.title.y = element_text(size = rel(1.6), angle = 90)) +
  theme(axis.title.x = element_text(size = rel(1.6), angle = 00))

ggplot(Glycogen_sel, aes(x = Stress, y = `mg/g`, colour = Feed))+
  geom_jitter()+
  ggtitle("Corrected Glycogen") +
  labs(x = "Stress", y = "Glycogen mg/g") +
  theme(axis.title.y = element_text(size = rel(1.6), angle = 90)) +
  theme(axis.title.x = element_text(size = rel(1.6), angle = 00))
```

#### Modelling of data (only corrected glycogen)

```{r Glycogen_modelling}
lm.Gly1 <- lm(`mg/g` ~ Treatment.Group, data = Glycogen_sel)
summary(lm.Gly1)
emmeans(lm.Gly1, pairwise~Treatment.Group)
plot(lm.Gly1, which =1)
```

#### Modelling Results
1. Stress had no impact on the level of glycogen when feeding was ad lib.  
2. Stress had a significant (P = 0.003) impact on the level of glycogen when feed was restricted.  
3. The combination of restricted feed and stress had the most significant (P < 0.001) effect on glycogen level, producing the lowest amount of glycogen.

## Glycogen Biopsy Data (Taken from M.semimembranosus Muscle)

```{r Glycogen_bios}
Glycogen_biopsy <- read_xlsx("Data/Copy of 20180413_DataSchool_Glycogen.xlsx", sheet = "Glycogen biopsy")
Glycogen_biopsy
```

####Tidying the data

```{r Glycogen_biops_tidying}
Glycogen_biopsy <- Glycogen_biopsy[1:4]
Glycogen_biopsy <- Glycogen_biopsy[-c(12:14),]
Glycogen_biopsy$Treatment.Group <- c(1,1,1,1,1,2,2,2,2,2,2)
Glycogen_biopsy_long <- Glycogen_biopsy %>% 
  gather(key = "Measure", value = "mg/g", "Glycogen mg/g", "Lactate mg/g", "Corrected Glycogen")
names(Glycogen_biopsy_long) <- c("ID", "Treatment.Group", "Measure", "mg/g")
Glycogen_biopsy_long[1:3] <- lapply(Glycogen_biopsy_long[1:3], as.factor)
Glycogen_biopsy_sel <- filter(Glycogen_biopsy_long, Measure == "Corrected Glycogen")
Glycogen_biopsy_long
```

#### Initial visulisation of data

```{r Glygogen_Biops_plot1, echo=FALSE}
ggplot(Glycogen_biopsy_sel, aes(x = Treatment.Group, y = `mg/g`, colour = Measure))+
  geom_point()+
  labs(x = "Treatment Group", y = "Glycogen mg/g") +
  theme(axis.title.y = element_text(size = rel(1.6), angle = 90)) +
  theme(axis.title.x = element_text(size = rel(1.6), angle = 00))
```

#### Modelling of data (only corrected glycogen)

```{r Glycogen_Biops_modelling}
lm.Glybiops1 <- lm(`mg/g` ~ Treatment.Group, data = Glycogen_biopsy_sel)
summary(lm.Glybiops1)
plot(lm.Glybiops1, which =1)
```

#### Modelling Results
1. Stress caused a significant (P = 0.0046) increase in M.semimembranosus muscle glycogen levels.

## Weight Gain Data

####Inital look at data
```{r Weight}
Weight_gain <- read_xlsx("Data/Copy of 20180413_DataSchool_Glycogen.xlsx", sheet = "Weight gain")
Weight_gain
```

####Tidying the data

```{r Weight_tidying}
Weight_gain <- Weight_gain[1:8]
Weight_gain <- Weight_gain[-c(61),]
names(Weight_gain) <- c("Treatment.Group", "ID", "-1", "0", "1", "2", "3", "4")
Weight_gain <- mutate(Weight_gain, Feed = ifelse(Treatment.Group <= 2, TRUE, FALSE)) %>%  #added feed column
  mutate(Stress = ifelse(Treatment.Group == 2|Treatment.Group == 4, TRUE, FALSE))
Weight_gain_long <- Weight_gain %>% 
  gather(key = "Week", value = "Weight.kg", "-1", "0", "1", "2", "3", "4")
Weight_gain_long[1:5] <- lapply(Weight_gain_long[1:5], as.factor)
Weight_gain_long$Feed <- relevel(Weight_gain_long$Feed, ref= 'TRUE')
Weight_gain_long
```

#### Initial visulisation of data

```{r Weight_plot1, echo=FALSE}
ggplot(Weight_gain_long, aes(x = Week, y = Weight.kg, colour = Stress))+
  geom_boxplot()+
  facet_wrap(~Feed) +
  ggtitle("Feed") +
  labs(x = "Week", y = "Weight kg") +
  theme(axis.title.y = element_text(size = rel(1.6), angle = 90)) +
  theme(axis.title.x = element_text(size = rel(1.6), angle = 00))
```

#### Modelling of data

```{r Weight_modelling}
lm.Weight1 <- lm(Weight.kg ~ Week + Stress*Feed, data = Weight_gain_long)
summary(lm.Weight1)
#emmeans(lm.Weight1, pairwise ~ Week + Stress*Feed)
plot(lm.Weight1, which =1)
```

#### Modelling Results
1. Before birth (-1 week) and at birth (0 week) there is no statistical difference in the weights of the sheep between treatment groups, as is expected because the treatments have not started.  
2. By the week 1 measurement, the treatment 4 group has a statistically significantly (P = 0.0236) lower (Est 3.033 +- 1.33 kg) weight than the other 3 groups. Meaning that the restricted feeding regime as well as stress is causing a lower sheep weight.  
3. By week 2 all treatments have lost a statistically significant (P = 0.0412) amount of weight compared to the previous weeks. Again, Treatment group 4 has lost statistically significantly (P = 0.0130) more weight in week 2 than the other groups.  
4. By week 3 treatment groups 1 and 2 have gained a statistically significant amount of weight compared to treatment groups 3 and 4, but between groups 1 and 2, and 3 and 4 there is no significant difference.  
5. By week 4 the trends are the same as week 3.

## Cortisol Data

####Inital look at data
```{r Cortisol}
Cortisol <- read_xlsx("Data/Copy of 20180413_DataSchool_Glycogen.xlsx", sheet = "Cortisol")
Cortisol
```

####Tidying the data

```{r Cortisol_tidying}
Cortisol <- Cortisol[1:5] #removing all columns that aren't needed
names(Cortisol) <- c("Treatment.Group", "ID", "1", "2", "3")
Cortisol <- Cortisol[-c(13),] #removing datapoint 15 as was replaced by 64
Cortisol <- mutate(Cortisol, Feed = ifelse(Treatment.Group <= 2, TRUE, FALSE)) %>%  #added feed column
  mutate(Stress = ifelse(Treatment.Group == 2|Treatment.Group == 4, TRUE, FALSE))
Cortisol_long <- Cortisol %>% 
  gather(key = "Time.point", value = "Cortisol_nmol/L", "1", "2", "3")
Cortisol_long[1:5] <- lapply(Cortisol_long[1:5], as.factor)
Cortisol_long$Feed <- relevel(Cortisol_long$Feed, ref= 'TRUE')
Cortisol_long
```

#### Initial visulisation of data

```{r Cortisol_plot1, echo=FALSE}
ggplot(Cortisol_long, aes(x = Treatment.Group, y = `Cortisol_nmol/L`, colour = Time.point))+
  geom_boxplot() +
  scale_y_log10()+
  ggtitle("Cortisol Concentration") +
  labs(x = "Treatment Group", y = "Cortisol nmol/L") +
  theme(axis.title.y = element_text(size = rel(1.6), angle = 90)) +
  theme(axis.title.x = element_text(size = rel(1.6), angle = 00))
```

#### Modelling of data

```{r Cortisol_modelling}
lm_cort <- lm(`Cortisol_nmol/L` ~Feed*Stress*Time.point, data = Cortisol_long)
anova(lm_cort)
summary(lm_cort)
plot(lm_cort, which =1)
```

#### Modelling Results

1. At time point 1, before the stress event, feed made no difference to the concentration of cortisol, but the non-stressed groups (1 & 3), contained a significantly (P = 0.003) higher cortisol concentration. This is strange because no stress event had occurred.  
2. There were no differences between treatment groups at time point after the stress event, therefore feed and stress makes no differnece to cortisol levels.  
3. Cortisol levels were significantly differnet at timepoint 3, post killing for all treatment groups.  
4. Restricted feed significantly (P = 0.004) lowered cortisol levels post killing.  
5. Stress significantly (P = 0.006) increased cortisol levels post killing.

## Peptide Matrix Data

####Data are the peak areas of identified peptides

```{r peptide}
peptide <- read_xlsx("Data/Copy of Data Matrix_Peptide Peak Area.xlsx")
peptide
```

####Tidy the data

```{r peptide_T}
peptide_T <- t(peptide) #transpose
peptide_T <- rownames_to_column(data.frame(peptide_T), 'ID')
peptide_T$ID[1] <- "ID"
peptide_T[] <- lapply(peptide_T, as.character)
peptide_T$X2[1] <- "Nutrition"
peptide_T$X1[1] <- "Treatment.Group"
colnames(peptide_T) <- peptide_T[1, ]
peptide_T <- peptide_T[-1 ,]
peptide_T[4:2117] <- lapply(peptide_T[4:2117], as.numeric)
peptide_T[1:3] <- lapply(peptide_T[1:3], as.factor)
peptide_T <- mutate(peptide_T, Stress = ifelse(Treatment.Group == 2|Treatment.Group == 4, TRUE, FALSE))
peptide_T <- select(peptide_T, Stress, everything())
peptide_T$Stress <- factor(peptide_T$Stress)
```
 
####Principle Component Analysis of the Data

```{r PCA}
pca_pep <- prcomp(peptide_T[5:2114], scale=TRUE)
```
###Plotting the PCA and Colouring by Qualitative Variables

```{r PCA_plots, echo=FALSE}
fviz_pca_ind(pca_pep, 
             habillage = peptide_T[1:48, 3]) +
  labs(title ="PCA Coloured by Groups", x = "Dim1 (30.9%)", y = "Dim2 (10 %)") +
  theme_minimal()

###Nutrition
fviz_pca_ind(pca_pep, 
             habillage = peptide_T[1:48, 1]) +
  labs(title ="PCA Coloured by Stress", x = "Dim1 (30.9%)", y = "Dim2 (10 %)") +
  theme_minimal()

###Treatment Group
fviz_pca_ind(pca_pep, 
             habillage = peptide_T[1:48, 4]) +
    labs(title ="PCA Coloured by Nutrition", x = "Dim1 (30.9%)", y = "Dim2 (10 %)") +
  theme_minimal()
```

####From the PCA plots, stress does not influence the abundance of individual peptides from the saliva, as expected, because the saliva was taken before the stress event.  Nutrition is having an effect on saliva peptide abundance though.  

###Plotting with the Nutrition Groups Coloured and with a 95% Confidence Ellipse

```{r PCA_plot2, echo=FALSE}
fviz_pca_ind(pca_pep, 
             habillage = peptide_T[1:48, 4], addEllipses = TRUE, 
             ellipse.level = 0.95, 
             title = "Saliva Peptide Analysis",
             repel = TRUE)+
  theme_minimal()

fviz_pca_var(pca_pep, select.var = list(contrib = 5), geom = c("point", "text"), repel = TRUE)

```

###The nutrition groups separate relatively well.  
###The top 5 peptides contributing the most influence on the PCA are:   
GTLDPVEK +2  
DAGTIAGLNVLR +2  
NSLESYAFNMK +2  
FIIPNIVK +2  
GNPTVEDLFTAK +2

